Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-07-30 Paper/assa.tex
comments ...

\section{Active subspaces and sensitivity analysis}
\label{sec:assa}
In this section, we provide key elements of the active subspace methodology
and provide theoretical connections between the global
sensitivity measures (DGSMs), activity scores, and the total Sobol' indices.

\subsection{Active subspaces}
\label{sub:ac}
As mentioned earlier, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output
due to uncertain inputs is predominantly captured along these directions.
Mathematically, these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows:
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where $\Omega$:~$[-1,1]^{N_p}$ is the domain of $\bm{\xi}$, and $N_p$ is the
number of uncertain inputs. \alennote{Why do we limit to $[-1, 1]^\Np$?} 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the considered parameter domain. 
%Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \bm{W}\bm{\Lambda}\bm{W}^\top
\ee
%
where $\bm{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdot \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \bm{W} = [\bm{w}_1~\bm{w}_2],~~\bm{\Lambda} = \begin{bmatrix}\bm{\Lambda}_1 & \\  &
  \bm{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\bm{w}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\bm{\Lambda}_1$ is the corresponding set of
eigenvalues. The active subspace is used to construct the so-called active
variables, $\bm{w}_1^\top \bm{\xi}$ and $f(\bm{\xi}(\theta))$ is transformed as
$G(\bm{w}_1^\top \bm{\xi})$. In practice, the set of samples, $\bm{\xi}_i$'s
are projected to the physical space as $\bm{\theta}_i$'s at which the model
output, $f(\bm{\xi}(\theta))$ is computed.  The matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\label{eq:chat}
 \ee
 %
%%It has been shown that the accuracy of approximated dominant eigenspace,
%%$\hat{\bm{w}}_1$ is inversely proportional to the difference between the
%%smallest eigenvalue in $\hat{\bm{\Lambda}}_1$ and the largest eigenvalue in
%%$\hat{\bm{\Lambda}}_2$~\cite{Constantine:2014}. 
%%Components of the eigenvectors in the active subspace could be used for
%%estimating the so-called activity scores as a measure for global sensitivity
%%and also be used for approximating the DGSMs as discussed in the following
%%section.

>>>>

\section{Active subspaces and sensitivity analysis}
\label{sec:assa}
%In this section, we provide key elements of the active subspace methodology and
%provide theoretical connections between the global sensitivity measures
%(DGSMs), activity scores, and the total Sobol' indices.

\subsection{Active subspaces}
\label{sub:ac}
As mentioned in the introduction, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output
due to uncertain inputs is predominantly captured along these directions.
Mathematically, these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows: (\alennote{Need to 
say what is the function $f$, first})
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where $\Omega$:~$[-1,1]^{N_p}$ is the domain of $\bm{\xi}$, and $N_p$ is the
number of uncertain inputs. \alennote{Why do we limit to $[-1, 1]^\Np$?} 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the considered parameter domain. 
%Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \bm{W}\bm{\Lambda}\bm{W}^\top
\ee
%
where $\bm{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \bm{W} = [\bm{w}_1~\bm{w}_2],~~\bm{\Lambda} = \begin{bmatrix}\bm{\Lambda}_1 & \\  &
  \bm{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\bm{w}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\bm{\Lambda}_1$ is the corresponding set of
eigenvalues. 
\alennote{$\vec{w}_1$ is a supposedly the matrix whose columns are 
the dominant eigenvectors. Denoting a matrix by a notation typically 
used for a vector is confusing. Use $\mat{W}_1$ instead?}

\alennote{The rest of this paragraph needs to be rewritten, as it is hard to 
see what is going on; what is $\theta$? I suggest, mentioning $\vec\theta$ 
and $\vec\xi$ earlier, and from then on make everything function of $\vec\xi$.
Explain the remaining ideas more clearly.}
The active subspace is used to construct the so-called active
variables, $\bm{w}_1^\top \bm{\xi}$ and $f(\bm{\xi}(\theta))$ is transformed as
$G(\bm{w}_1^\top \bm{\xi})$. In practice, the set of samples, $\bm{\xi}_i$'s
are projected to the physical space as $\bm{\theta}_i$'s at which the model
output, $f(\bm{\xi}(\theta))$ is computed.  The matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\label{eq:chat}
 \ee
 %
%%It has been shown that the accuracy of approximated dominant eigenspace,
%%$\hat{\bm{w}}_1$ is inversely proportional to the difference between the
%%smallest eigenvalue in $\hat{\bm{\Lambda}}_1$ and the largest eigenvalue in
%%$\hat{\bm{\Lambda}}_2$~\cite{Constantine:2014}. 
%%Components of the eigenvectors in the active subspace could be used for
%%estimating the so-called activity scores as a measure for global sensitivity
%%and also be used for approximating the DGSMs as discussed in the following
%%section.

### 2018-07-30 Paper/intro.tex
comments ...

Time evolution of a chemically reacting system is largely dependent upon rate constants
associated with individual reactions. The rate constants are typically assumed to exhibit
a certain correlation with temperature (e.g. Arrhenius-type). Hence, accurate specification
of the rate-controlling parameters is critical to the fidelity of simulations. However,
in practical applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data. Consequently, the
underlying uncertainty associated with the parameters and therefore the simulation-prediction
is typically ignored. Vohra et al. calibrated the Arrhenius diffusion parameters to quantify
the uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian framework
in~\cite{Vohra:2014}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments. Additionally, 
in~\cite{Vohra:2017}, a design strategy was implemented to determine optimal conditions
for experiments in order to enhance the information gain in the uncertain diffusion parameters.
Bayesian calibration was accelerated in these efforts using a polynomial 
chaos~\cite{Ghanem:2003, Xiu:2002} surrogate for the model. However,
the number of uncertain parameters for calibration were limited to two in both cases. In
higher dimensions, the construction of surrogate as well as the calibration process become
remarkably challenging due to increase in computational effort.

In this paper, we focus our efforts on mitigating the computational challenge posed by uncertainty
quantification (UQ) for applications involving a large number of chemical reactions. Specifically, we
accomplish this by implementing an efficient approach that aims to reduce the dimensionality
of the problem by computing the so-called active subspace~\cite{Constantine:2015}. 
The active subspace, discussed later in~\ref{sub:ac}, essentially constitutes the dominant
eigenspace of a matrix derived from the gradient information of the model output with respect
to an input. The dominant eigenspace essentially comprises `important' directions in the parameter
space that predominantly capture the variability in the model output due to uncertainty in the
inputs.

The high-dimensional application problem considered in this work is the H$_2$/O$_2$
reaction mechanism~\cite{Yetter:1991}. This mechanism has been gaining a lot of attention in the
recent years as a potential source of clean energy for locomotive applications~\cite{Das:1996}.   
The mechanism involves 19 reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in table~\ref{tab:kinetics}, 
the reaction rate is assumed to show an Arrhenius correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
%
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
%
The equivalence ration ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer (O$_2$) ratio at given
condition to the same quantity under stoichiometric conditions. In this study, computations were performed
at fuel-rich conditions, $\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of temperature increase
to exceed a given threshold, regarded as `ignition delay' is recorded.

We are interested in investigating the impact of uncertainty in the rate-controlling parameters, 
pre-exponent~($A_i$) and the activation energy~($E_{a,i}$) on ignition delay. The $A_i$'s are
considered to be uniformly distributed about their nominal estimates provided in~\cite{Yetter:1991}.
The total number of uncertain inputs is 33 which makes the
present problem reasonably high-dimensional involving a significantly large amount of computational
effort. To reduce the effort, we explore two strategies: one involving the use of gradients (grad-based)
and the other involving local linear approximation which does require gradient computation (grad-free).
These strategies are based on algorithms 1.1 and 1.2 provided  in~\cite{Constantine:2015}. However, it
must be noted that in this work, an iterative approach is adopted for implementing the two strategies.
Consequently, the amount of computational effort appears to have decreased, and is found to be
relatively less in the case of grad-free approach as discussed later in section~\ref{sec:method}.

Components of the dominant eigenvectors can be used to estimate the so-called 
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into the relative importance
of the uncertain parameters. In 
section~\ref{sub:gsa}, we demonstrate that the activity scores could be used to approximate the 
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and establish a relationship between
the Sobol' total effect index, the DGSMs, and the activity scores. Below, we summarize key contributions of
the paper and its organization.

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity scores as means for approximating
the DGSMs is provided. (2) A general relationship between DGSMs, activity scores, and the total Sobol' index 
is established. (3) Iterative gradient-based and gradient-free approaches for computing the active subspace are
 demonstrated. The two approaches are shown to be consistent and yield greater
computational gains compared to the 
 conventional approach. As mentioned above, the gains are greater in the case of grad-free approach. (4) The
 grad-free approach is implemented for discovering an active subspace in a 33-dimensional parameter space 
 for the H$_2$/O$_2$ reaction kinetics application.

>>>>

Time evolution of a chemically reacting system is largely dependent upon rate
constants associated with individual reactions. The rate constants are
typically assumed to exhibit a certain correlation with temperature (e.g.
Arrhenius-type). Hence, accurate specification of the rate-controlling
parameters is critical to the fidelity of simulations. However, in practical
applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data.
(\alennote{Any references for this?})
Consequently, the underlying uncertainty associated with the parameters and
therefore the simulation is typically ignored.

Vohra et al.~calibrated the Arrhenius diffusion parameters to quantify the
uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian
framework in~\cite{Vohra:2014}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments.
Additionally, in~\cite{Vohra:2017}, a design strategy was implemented to
determine optimal conditions for experiments in order to enhance the
information gain in the uncertain diffusion parameters.  Bayesian calibration
was accelerated in these efforts using a polynomial chaos~\cite{Ghanem:2003,
Xiu:2002} surrogate for the model. However, the number of uncertain parameters
for calibration were limited to two in both cases. In higher dimensions, the
construction of surrogate as well as the calibration process become remarkably
challenging due to increase in computational effort.~\alennote{I don't see why
these references are being cited here. Are these cited as examples of efforts
in calibrating chemical systems?}

In this paper, we focus on mitigating the computational challenge
posed by uncertainty quantification (UQ) for applications involving a large
number of chemical reactions. We accomplish this by implementing
an efficient approach that aims to reduce the dimensionality of the problem. This is
done by
computing the so-called active subspace~\cite{Constantine:2015}.  The active
subspace, discussed later in~\ref{sub:ac}, constitutes the dominant
eigenspace of a matrix derived from the gradient information of the model
output with respect to an input. The dominant eigenspace essentially comprises
``important'' directions in the parameter space that predominantly capture the
variability in the model output due to uncertainty in the inputs.

The application problem considered in this work is the
H$_2$/O$_2$ reaction mechanism~\cite{Yetter:1991}. This mechanism has been
gaining a lot of attention in the recent years as a potential source of clean
energy for locomotive applications~\cite{Das:1996}. (\alennote{you say recent years, but
the reference is from 1996, are there more recent papers you can refer to?}.   
The mechanism involves 19
reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in
table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
The equivalence ratio ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer
(O$_2$) ratio at given condition to the same quantity under stoichiometric
conditions. In this study, computations were performed at fuel-rich conditions,
$\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of
temperature increase to exceed a given threshold, regarded as \emph{ignition delay}
is recorded.

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging. 
To reduce the effort, we explore two
strategies: 
a gradient-based approach and  
a gradient-free approach.
%involving local linear approximation which does notrequire gradient computation
%(grad-free).  
\alennote{Thorough the paper 
change grad-based to gradient-based, and similarly for grad-free}. 
%
These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. In this work, an
iterative approach is adopted for implementing the two strategies.
Consequently, the amount of computational effort appears to have decreased
\alennote{This will be unclear to readers}, and
is found to be relatively less in the case of grad-free approach as discussed
later in section~\ref{sec:method}.

\alennote{This paragraph will not be meaningful to readers, especially those not familiar
with active subspaces; Need to explain DGSMs, their importance, 
and say something general about our
approach for approximating them.}
Components of the dominant eigenvectors can be used to estimate the so-called
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into
the relative importance of the uncertain parameters. In section~\ref{sub:gsa},
we demonstrate that the activity scores could be used to approximate the
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and
establish a relationship between the Sobol' total effect index, the DGSMs, and
the activity scores.

%Below, we summarize key contributions of the paper and its
%organization.

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity
scores as means for approximating the DGSMs is provided. (2) A general
relationship between DGSMs, activity scores, and the total Sobol' index is
established. (3) Iterative gradient-based and gradient-free approaches for
computing the active subspace are demonstrated. The two approaches are shown to
be consistent and yield greater computational gains compared to the
conventional approach. As mentioned above, the gains are greater in the case of
grad-free approach. (4) The grad-free approach is implemented for discovering
an active subspace in a 33-dimensional parameter space for the H$_2$/O$_2$
reaction kinetics application. 
\alennote{This says too much. Focus on the actual novel 
contributions; also, the application should be made more prominent.}
